---
title: Countries with High External Health Funding
format: html
---

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(nanoparquet)

country <- read_parquet("country.parquet")
country_year <- read_parquet("country-year.parquet")
health_spending <- read_parquet("health_spending.parquet")
```

## Which countries rely most on external funding?

```{r}
#| label: top-ext-countries
ext <- health_spending |>
  filter(expenditure_type == "ext") |>
  left_join(country, by = "iso3_code")

# Average external share of CHE over the most recent 5 years
ext_recent <- ext |>
  filter(year >= 2019) |>
  summarise(
    mean_che = mean(che, na.rm = TRUE),
    .by = c(iso3_code, country_name)
  ) |>
  arrange(desc(mean_che))

ext_recent |> head(20)
```

The countries with the highest external health funding share have several things in common:

- **Sub-Saharan Africa dominates.** The majority are African countries, reflecting the continent's heavy reliance on development aid for health systems.
- **Small island developing states (SIDS).** A second cluster is tiny Pacific and Caribbean island nations (Micronesia, Marshall Islands, Niue, Tonga, Vanuatu, etc.).
  Their small populations mean even modest aid translates to a huge share of total health spending.
- **Fragile and conflict-affected states.** Several are current or recent conflict zones (South Sudan, Somalia, Haiti, CAR, DRC, Burundi), where weak domestic revenue means external donors fill the gap.
- **Very small total health budgets.** Almost all have total health spending under $1--2B.
  The exceptions (Ethiopia, Tanzania, DRC, Uganda) have very large populations, so per-capita spending is still extremely low.

Focus on the top 20 countries by average external share (2019–2023).

```{r}
#| label: top-countries
top20 <- ext_recent |> slice_head(n = 20)
```

## External funding share over time

```{r}
#| label: fig-ext-share-over-time
#| fig-width: 10
#| fig-height: 8
ext |>
  semi_join(top20, by = "iso3_code") |>
  mutate(
    country_name = fct_reorder(country_name, -che, .fun = mean, na.rm = TRUE)
  ) |>
  ggplot(aes(year, che)) +
  geom_line() +
  facet_wrap(~country_name, ncol = 4) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    x = NULL,
    y = "External funding (% of CHE)",
    title = "External health funding share, top 20 countries"
  )
```

## External funding in absolute terms

```{r}
#| label: fig-ext-absolute
#| fig-width: 10
#| fig-height: 8
ext |>
  semi_join(top20, by = "iso3_code") |>
  mutate(
    country_name = fct_reorder(
      country_name,
      -usd2023,
      .fun = mean,
      na.rm = TRUE
    )
  ) |>
  ggplot(aes(year, usd2023)) +
  geom_line() +
  facet_wrap(~country_name, ncol = 4, scales = "free_y") +
  scale_y_continuous(
    labels = scales::label_dollar(scale = 1e-6, suffix = "M")
  ) +
  labs(
    x = NULL,
    y = "External funding (2023 USD)",
    title = "External health funding, top 20 countries"
  )
```

## Per capita health spending

```{r}
#| label: fig-per-capita
#| fig-width: 10
#| fig-height: 8
che_pc <- health_spending |>
  filter(expenditure_type == "che") |>
  left_join(country_year, by = c("iso3_code", "year")) |>
  mutate(usd_per_capita = usd2023 / population) |>
  semi_join(top20, by = "iso3_code") |>
  left_join(country, by = "iso3_code") |>
  mutate(
    country_name = fct_reorder(
      country_name,
      -usd_per_capita,
      .fun = mean,
      na.rm = TRUE
    )
  )

ggplot(che_pc, aes(year, usd_per_capita)) +
  geom_line() +
  facet_wrap(~country_name, ncol = 4, scales = "free_y") +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(
    x = NULL,
    y = "Health spending per capita (2023 USD)",
    title = "Per capita health expenditure, top 20 externally funded countries"
  )
```

## How does external funding relate to total spending?

```{r}
#| label: fig-ext-vs-che
#| fig-width: 10
#| fig-height: 6
che <- health_spending |>
  filter(expenditure_type == "che", year >= 2019) |>
  summarise(mean_che_usd = mean(usd2023), .by = iso3_code)

ext_vs_che <- ext_recent |>
  left_join(che, by = "iso3_code")

ggplot(ext_vs_che, aes(mean_che_usd, mean_che)) +
  geom_point(alpha = 0.5) +
  ggrepel::geom_text_repel(
    data = ext_vs_che |> filter(mean_che > 30),
    aes(label = country_name),
    size = 3
  ) +
  scale_x_log10(labels = scales::label_dollar(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    x = "Total health expenditure (2023 USD, log scale)",
    y = "External funding (% of CHE)",
    title = "External funding share vs total health spending (2019–2023 avg)"
  )
```

## Funding source breakdown for high-external countries

```{r}
#| label: fig-funding-breakdown
#| fig-width: 10
#| fig-height: 8
hs_labels <- c(
  gghed = "Government",
  pvtd = "Private",
  ext = "External"
)

health_spending |>
  filter(expenditure_type != "che") |>
  semi_join(top20, by = "iso3_code") |>
  left_join(country, by = "iso3_code") |>
  mutate(
    country_name = fct_reorder(
      country_name,
      -che * (expenditure_type == "ext"),
      .fun = mean,
      na.rm = TRUE
    )
  ) |>
  ggplot(aes(year, che, fill = expenditure_type)) +
  geom_area() +
  facet_wrap(~country_name, ncol = 4) +
  scale_fill_brewer(palette = "Set2", labels = hs_labels) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    x = NULL,
    y = "% of CHE",
    fill = NULL,
    title = "Health funding sources, top 20 externally funded countries"
  ) +
  theme(legend.position = "bottom")
```
